---
title: "NBA Over Time Analysis"
author: "Jack Kincannon"
date: "12/3/2020"
output: pdf_document
---

# NBA Shot Analysis 2009-10 vs 2018-19

## Load Packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(dplyr)
library(here)
library(patchwork)
library(ggplot2)
library(RColorBrewer)
library(cowplot)
library(magick)
library(png)
theme_set(theme_cowplot())
```

## Shot Distributions

```{r}
shots_2010 <- read.csv("data/nba_shots_2010.csv")
#glimpse(shots_2010)
```

```{r}
x1 <- c("Mid-Range", "Right Corner 3", "Left Corner 3", "Above the Break 3", "Backcourt")
mypng <- readPNG("/home/jdkincan/1001-proj-final/nba_court_1.png")

dat3 <- shots_2010 %>%
  filter(SHOT_ZONE_BASIC %in% x1) 

plot2 <- ggplot(dat3, aes(LOC_X, LOC_Y)) + 
  geom_hex(binwidth = 20, na.rm = FALSE) +
  scale_fill_gradient(low="midnightblue", high="brown1") +
  labs(title = "All Shots Taken Outside of the Paint", subtitle = "2009-2010 NBA Season", x = "", y = "")
plot2

#ggdraw() +
  #draw_image(mypng) +
  #draw_plot(plot2)
```

```{r, message = FALSE}
shots_2019 <- read_csv("data/nba_shots_2019.csv")
#glimpse(shots_2019)
```

```{r, echo=FALSE}
x <- c("Mid-Range", "Right Corner 3", "Left Corner 3", "Above the Break 3", "Backcourt")

dat2 <- shots_2019 %>%
  filter(SHOT_ZONE_BASIC %in% x)

plot1 <- ggplot(dat2, aes(LOC_X, LOC_Y)) + 
  geom_hex(binwidth = 20, na.rm = FALSE) +
  scale_fill_gradient(low="midnightblue", high="brown1") +
  labs(title = "All Shots Taken Outside of the Paint", subtitle = "2018-2019 NBA Season", x = "", y = "")

plot1

glimpse(shots10_pct)
```

These plots show how the NBA has evolved in just 10 years. The lack of mid-rangers is even more apparent when looking at data from years before 2010. However, my method of collection only allowed this range. 

## Shot Percentages and Points Per Shot

```{r}


# 2010 shots
dat13 <- dat9 %>%
  full_join(dat10, by = "SHOT_DISTANCE")

# 2019 shots
dat14 <- dat11 %>%
  full_join(dat12, by = "SHOT_DISTANCE")

# do 4 graphs 2 and 3 pointers 


ggplot(economics, aes(x=date)) + 
  geom_line(aes(y = psavert), color = "darkred") + 
  geom_line(aes(y = uempmed), color="steelblue", linetype="twodash") 

plot7 <- ggplot(shots10_pct, aes(x = SHOT_DISTANCE)) +
  geom_line(aes(y = points_per_shot.x), color = "skyblue1") +
  geom_line(aes(y = points_per_shot.y), color = "deepskyblue1") +
  labs(x = "Shot Distance (ft)", y = "Points Per Shot") +
  theme(legend.position = "right") +
  geom_vline(xintercept = 22, color = "firebrick", label = "Corner 3 Dist.") +
  geom_vline(xintercept = 23, color = "firebrick", label = "Top-of-Key 3 Dist.") +
  xlim(1, 34) +
  ylim(0.0, 1.5)
plot7



# calculte break even point for both years

p6 <- ggplot(dat13, aes(x = SHOT_DISTANCE, y = points_per_shot.y)) +
  geom_line() +
  labs(x = "Shot Distance (ft)", y = "") +
  geom_hline(yintercept = 1, color = "brown1") +
  geom_vline(xintercept = 26, color = "brown1") +
  xlim(24, 35) +
  ylim(0.0, 1.5)

p8 <- ggplot(dat14, aes(x = SHOT_DISTANCE, y = points_per_shot.x)) +
  geom_line() +
  labs(x = "Shot Distance (ft)", y = "Points Per Shot") +
  geom_hline(yintercept = 1, color = "brown1") +
  geom_vline(xintercept = 2, color = "brown1") +
  xlim(1, 23) +
  ylim(0.0, 1.5)


p10 <- ggplot(dat14, aes(x = SHOT_DISTANCE, y = points_per_shot.y)) +
  geom_line() +
  labs(x = "Shot Distance (ft)", y = "") +
  geom_hline(yintercept = 1, color = "brown1") +
  geom_vline(xintercept = 28, color = "brown1", linetype) +
  xlim(24, 35) +
  ylim(0.0, 1.5)

p11 <- (p3 + p5) / (p4 + p6)
p12 <- (p7 + p9) / (p8 + p10)

#interested in p4 p6 p8 p10

# p11 is 2010 Data
# p12 is 2019 Data

p11
p12
```
These graphs shows that is is much more valuable to shoot a high percentage 3 pointer than it is to shoot anything besides a layup inside the 3 point line. 

## BROKEN CHUNK

```{r}
dat3 <- shots_2010 %>%
  mutate(mean_dist_10 = sum(SHOT_DISTANCE) / length(SHOT_DISTANCE))
#str(dat3)

dat4 <- shots_2019 %>%
  mutate(mean_dist_19 = sum(SHOT_DISTANCE) / length(SHOT_DISTANCE)) 
#str(dat4)

N_19 <- length(shots_2019$SHOT_DISTANCE)
N_10 <- length(shots_2010$SHOT_DISTANCE)

dat5 <- dat3 %>%
  summarize(sd_shot_dist = round(sd(SHOT_DISTANCE), 4),
            se_shot_dist = round(sd(SHOT_DISTANCE) / sqrt(N_10), 4),
            mean_dist_10)

dat6 <- dat4 %>%
  summarize(sd_shot_dist = round(sd(SHOT_DISTANCE), 4),
            se_shot_dist = round(sd(SHOT_DISTANCE) / sqrt(N_19), 4),
            mean_dist_19) 

dat7 <- sample(dat5, size = 20000, replace = TRUE, prob = NULL)
dat8 <- sample(dat6, size = 20000, replace = TRUE, prob = NULL)

## shot_dist_t_test <- t.test(dat7, dat8)
```